---
title: 'Extensions to ggplot2 are easy – right?'
author:
  - Jeffrey Stevens
date: '2024-12-05'
categories:
  - extensions
  - geoms
  - ggplot2
  - talk
description: "Inaugural Nebraska R User Group talk given by Dr. Heike Hofmann."
draft: true
---

On December 4, [Dr. Heike Hofmann](https://github.com/heike) gave the inaugural Nebraska R User Group talk on "Extensions to ggplot2 are easy – right?" ([slides](https://heike.github.io/NebRUG-2024/2024-12-04-ggplot2-extensions/ggplot2-extensions.html)). Many of us have used the [`{ggplot2}`](https://ggplot2.tidyverse.org/) package to plot our data. But have you ever wanted to create new ways of plotting data that weren't available? `{ggplot2}` is extensible, letting you create your own custom plotting functions. Dr. Hofmann walked us through how to do this. Here are a few highlights.

# Setting up the data

First, we need a good data set to work with. Dr. Hofmann recommended check the [#TidyTuesday](https://github.com/rfordatascience/tidytuesday) website for data sets. She chose to use the [Southern Resident Killer Whale Encounters](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-10-15/readme.md) data set, which tracks the location of orcas in the Pacific Northwest’s Salish Sea by the [Center for Whale Research](https://www.whaleresearch.com/encounters). 

![Source: [Center for Whale Research](https://www.whaleresearch.com/2024-69)](images/orca_cwr.jpg){fig-alt="Black and white orca breaching out of the water"}

Let's import the data from the #TidyTuesday data repository.

```{r message=FALSE}
orcas <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-10-15/orcas.csv')
```

A quick plot of this data shows the locations of encounters:

```{r warning=FALSE}
library(ggplot2)
orcas |> 
  ggplot(aes(x = begin_longitude, y = begin_latitude)) +
  geom_point()
```

# Extending ggplot

But what if we want to draw a polygon around the perimeter of the data points to show the extent of the area covered by the data? In geometry, this is called a [convex hull](https://en.wikipedia.org/wiki/Convex_hull). There is not a function in `{ggplot2}` that automatically plots convex hulls. So we need to create our own extension to do this.

As a reminder, we use _geoms_ to define how data are mapped to physical properties of the plots (e.g., points, lines, polygons, text). So above we used the `geom_points()` function to map the longitude and latitude data to x- and y-coordinates on our Cartesian coordinate system.

But for every _geom_ there is a _stat_ that computes statistics on the data before mapping to the plot via the _geom_. The _stat_ underlying `geom_points()` is `stat_identity()`. The "identity" part just means that the raw data is passed directly to the _geom_ (there are no actual computations run on the data).

Therefore, to create a new type of plotting function, we must create a new _geom_ and a new _stat_. Fortunately, we don't have to do these from scratch. We can modify existing _geoms_ and _stats_. So it makes sense to find the existing _geom_ that is closest to what you want to do, and modify that.

To plot a convex hull, we must first find the hull, which means finding the points that represent the corners of the hull. So our new plotting function must find a subset of points and then plot a polygon connecting those points. 

# Creating _stat_ objects and functions

Before we create a new _stat_ function, we have to create a new _stat_ object that computes the convex hull. Fortunately, there is a function for that (`grDevices::chull()`) that we can embed in our _stat_ object. So here we use `ggproto()` to create a new object based on the prototype _stat_ that computes for each group the subset of data points that represent the convex hull.

```{r}
StatChull <- ggproto(
  "StatChull", Stat,
  required_aes = c("x", "y"),
  compute_group = function(data, scales) {
     data[chull(data$x, data$y), , drop = FALSE]
  }
)
```

Now we need to create the _stat_ function. Because we need the function to draw a polygon based on the convex hull points, we will use the `polygon` _geom_ as the base for this function. We include a `layer()` function and keep most of the rest of the arguments this same to avoid messing up all of the automatic benefits we get of using the `polygon` _geom_.

```{r}
stat_chull <- function(
    mapping = NULL, data = NULL, geom = "polygon",
    position = "identity", na.rm = FALSE,
    show.legend = NA,  inherit.aes = TRUE, ...) {
  layer(
    stat = StatChull, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, 
    inherit.aes = inherit.aes, params = list(na.rm = na.rm, ...)
  )
}
```

We can now use our new `stat_chull()` function two draw the convex hull around the data. 
```{r warning=FALSE}
orcas |> 
  ggplot(aes(x = begin_longitude, y = begin_latitude)) +
  stat_chull() + 
  geom_point() 
```

We have drawn a convex hull!  But the fill and transparency defaults aren't great. Fortunately, we can pass standard aesthetic arguments to the function to control line color and polygon fill and transparency.

```{r warning=FALSE}
orcas |> 
  ggplot(aes(x = begin_longitude, y = begin_latitude)) +
  stat_chull(fill="grey60", colour = "grey30", alpha = 0.8) + 
  geom_point() 
```
# Creating _geom_ objects and functions

But if we want to set different defaults from the original _geom_ (in this case `polygon`), we also need to set up our own _geom_ object and function. We will create a new _geom_ object called `GeomChull` based on the prototype object `GeomPolygon`. Then we'll define new aesthetic defaults.

```{r}
GeomChull <- ggproto(
  "GeomChull", GeomPolygon,
  default_aes = ggplot2::aes(
    colour = "grey30", fill = "grey50", alpha = 0.5, # new ones
    linewidth=0.5, linetype = 1, subgroup=NULL
    )
)
```

With the new _geom_ object in place, we need to create a new _geom_ function that uses the `chull` stat that we created and the `GeomChull` object.

```{r warning=FALSE}
geom_chull <- function (mapping = NULL, data = NULL, 
                        stat = "chull", position = "identity", 
    rule = "evenodd", ..., na.rm = FALSE, show.legend = NA, inherit.aes = TRUE) 
{
    layer(data = data, mapping = mapping, stat = stat, 
          geom = GeomChull, position = position, 
          show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(na.rm = na.rm, rule = rule, ...))
}
```

Now we can use `geom_chull()` instead of `stat_chull()` and we have a new set of default aesthetics.

```{r warning=FALSE}
orcas |> 
  ggplot(aes(x = begin_longitude, y = begin_latitude)) +
  geom_chull() + 
  geom_point() 
```

# Bonus side effects

Again, because we left so many of the arguments the same, we can do cool things like automatically have the function apply to facets.

```{r warning=FALSE}
orcas |> 
  ggplot(aes(x = begin_longitude, y = begin_latitude)) +
  geom_chull() + 
  geom_point() +
  facet_wrap(~lubridate::month(date))
```

Or apply the function to groups.

```{r warning=FALSE}
orcas |> 
  ggplot(aes(x = begin_longitude, y = begin_latitude, 
             colour = year, group = year)) +
  geom_chull() + 
  geom_point() 
```

# Other extensions

See---we told you extensions to ggplot2 would be easy – right? For other ggplot2 extensions that folks have created, check out the [ggplot2 extensions gallery](https://exts.ggplot2.tidyverse.org/gallery/), which includes extensions such as [`{ggmosaic}`](https://github.com/haleyjeppson/ggmosaic), [`{ggpcp}`](https://heike.github.io/ggpcp/), and [`{ggrepel}`](https://github.com/slowkow/ggrepel) just to name a few!


